Endogenous Macrodynamics in Algorithmic Recourse

### 2022-08-12 dev/paper/paper.tex

\begin{document}
\maketitle
\begin{abstract}
Existing work on Counterfactual Explanations (CE) and Algorithmic
Recourse (AR) has largely been limited to the static setting: given some
classifier we are interested in finding close, actionable, realistic,
sparse, diverse and ideally causally founded counterfactuals. The
ability of CE to handle dynamics like data and model drift remains a
largely unexplored research challenge at this point. Only one recent
work considers the implications of exogenous domain and model shifts.
This project instead focuses on endogenous dynamics, that is shifts that
occur when AR is actually implemented by a proportion of individuals.
Early findings suggest that the involved shifts may be large with
important implications on the validity of AR and the overall
characteristics of the sample population.
\end{abstract}
\ifdefined\Shaded\renewenvironment{Shaded}{\begin{tcolorbox}[interior hidden, breakable, borderline west={3pt}{0pt}{shadecolor}, enhanced, frame hidden, boxrule=0pt, sharp corners]}{\end{tcolorbox}}\fi

@@

Models or algorithms that fall into this category are typically referred
to \textbf{black-box} models. Despite their shortcomings, black-box
models have grown in popularity in recent years and have at times
created undesirable societal outcomes
\protect\hyperlink{ref-o2016weapons}{{[}2{]}}. The scientific community
has tackled this issue from two different angles: while some have
appealed for a strict focus on inherently iterpretable models
\protect\hyperlink{ref-rudin2019stop}{{[}3{]}}, others have investigated
different ways to explain the behaviour of black-box models. These two
sub-domains can be broadly referred to as \textbf{interpretable AI} and
\textbf{explainable AI} (XAI), respectively.

Among the approaches to XAI that have recently grown in popularity are
\textbf{Counterfactual Explanations} (CE). They explain how inputs into
a model need to change for it to produce different outputs.
Counterfactual Explanations that involve realistic and actionable
changes can be used for the purpose of \textbf{Algorithmic Recourse}
(AR) to help individuals who face adverse outcomes. An example relevant
to the Social Sciences is consumer credit: in this context AR can be
used to guide individuals in improving their creditworthiness, should
they have previously been denied access to credit based on an automated
decision-making system. A meaningful recourse recommendation for a
denied applicant could be: \emph{``If your net savings rate had been
10\% of your monthly income instead of the actual 8\%, your application
would have been successful. See if you can temporarily cut down on
consumption.''} In the remainder of this paper we will use both
terminologies - recourse and counterfactual - interchangeably to refer
to situations where counterfactuals are generated with the intent to
provide individual recourse.

Existing work in this field has largely worked in a static setting:
various approaches have been proposed to generate counterfactuals for a
given individual that is subject to some pre-trained model. More recent
work has compared different approaches within this static setting
\protect\hyperlink{ref-pawelczyk2021carla}{{[}4{]}}. In this work we go
one step further and ask ourselves: what happens if recourse is provided
and implemented repeatedly? What types of dynamics are introduced and
how do different counterfactual generators compare in this context?

Research on algorithmic recourse has also so far typically addressed the
issue from the perspective of one single individual and has indeed been
referred to as \textbf{individual recourse} in some places. Arguably
though, most real-world applications that warrant algorithmic recourse
involve potentially large groups of individuals typically competing for
scarce resources. Our work demonstrates that in such scenarios, choices
made by or for one single individual are likely to affect the broader
collective of individuals in ways that current approaches to AR fail to
account for. More specifically, we argue that a strict focus on
minimizing the private costs faced by individuals may be too narrow an
objective.

@@

To the best of our knowledge this is the first work investigating
endogenous macrodynamics in AR. Our contributions to the state of
knowledge are as follows: firstly, we posit a compelling argument that
calls for a novel perspective on algorithmic recourse extending our
focus from single individuals to groups. Secondly, we introduce an
experimental framework extending previous work by
\protect\hyperlink{ref-altmeyer2022CounterfactualExplanations}{{[}6{]}},
which enables us to study macrodynamics of algorithmic recourse through
simulations that can be fully parallelized. Thirdly, we use this
framework to provide a first in-depth analysis of endogenous recourse
dynamics induced by various popular counterfactual generators including
\protect\hyperlink{ref-wachter2017counterfactual}{{[}5{]}},
\protect\hyperlink{ref-schut2021generating}{{[}7{]}},
\protect\hyperlink{ref-joshi2019towards}{{[}8{]}},
\protect\hyperlink{ref-mothilal2020explaining}{{[}9{]}} and
\protect\hyperlink{ref-antoran2020getting}{{[}10{]}}. To this end we
propose a number of novel evaluation metrics that can be used to
quantify and benchmark the macrodynamics introduced by the different
generators. Finally, we also discuss what drives endogenous dynamics and
propose strategies to mitigate them.

The remainder of the paper is structured as follows:
Section~\ref{sec-related} places our work in the broader context of
related literature. Section~\ref{sec-method} presents our methodology
and data. Section~\ref{sec-empirical} presents our empirical findings
which are then discussed in the broader context of the literature in
Section~\ref{sec-discussion}. We also point to some of the limitations
or our work as well as avenues for future research in
Section~\ref{sec-limit}. Finally, Section~\ref{sec-conclusion}
concludes.

@@

In this Section we provide a review of the relevant literature. First,
Section~\ref{sec-related-recourse} discusses the existing research
within the domain of counterfactual explanations and algorithmic
recourse. Then, Section~\ref{sec-related-shifts} presents some of the
previous work on the measurement of dataset and model shifts.

@@

A framework for Counterfactual Explanations was first proposed in 2017
by \protect\hyperlink{ref-wachter2017counterfactual}{{[}5{]}} and has
served as the baseline for most methodologies that have been proposed
since then. Let \(M: \mathcal{X} \mapsto \mathcal{Y}\) denote some
pre-trained model that maps from inputs \(X \in \mathcal{X}\) to outputs
\(Y \in \mathcal{Y}\). Then we are interested in minimizing the
complexity or effort \(H=h(x\prime)\) associated with moving an
individual \(x\) to a counterfactual state \(x\prime\) such that the
predicted outcome \(M(x\prime)\) corresponds to some target outcome
\(t\):

\begin{equation}\protect\hypertarget{eq-obj}{}{
\min_{x\prime \in \mathcal{x}} c(x\prime) \ \ \ \mbox{s. t.} \ \ \ M(x\prime) = t
}\label{eq-obj}\end{equation}

@@

\begin{equation}\protect\hypertarget{eq-solution}{}{
x\prime = \arg \min_{x\prime}  \ell(M(x\prime),t) + \lambda h(x\prime)
}\label{eq-solution}\end{equation}

In the baseline work and many subsequent approaches the complexity
function \(h: \mathcal{X} \mapsto \mathbb{R}\) is proxied by some
distance metric based on the simple intuition that large perturbations
of \(x\) are costly.

Many approaches for the generation of algorithmic recourse have been
described in the literature. An October 2020 survey by Karimi et
al.~layed out 60 algorithms that have been proposed since 2014
\protect\hyperlink{ref-karimi2020survey}{{[}11{]}}. Another survey
published around the same time by Verma et al.~described 29 algorithms
\protect\hyperlink{ref-verma2020counterfactual}{{[}12{]}}. Different
approaches vary primarily in terms of the objective functions they
impose, how they optimize said objective (from brute force through
gradient-based approaches to graph traversal algorithms), and how the
ensure that certain requirements for CE are met. Regarding the latter,
the literature has produced an extensive list of desiderata each
addressing different needs. To name but a few, we are interested in
generating counterfactuals that close
\protect\hyperlink{ref-wachter2017counterfactual}{{[}5{]}}, actionable
\protect\hyperlink{ref-ustun2019actionable}{{[}13{]}}, realistic
\protect\hyperlink{ref-schut2021generating}{{[}7{]}}, sparse, diverse
\protect\hyperlink{ref-mothilal2020explaining}{{[}9{]}} and if possible
causally founded
\protect\hyperlink{ref-karimi2021algorithmic}{{[}14{]}}.

Efforts so far have largely been directed at improving the quality of
counterfactual explanations within a static context: given some
pre-trained classifier \(M: \mathcal{X} \mapsto \mathcal{Y}\) we are
interested in generating one or multiple meaningful counterfactual
explanations for some individual characterized by \(x_i\). The ability
of counterfactual explanations to handle dynamics like data and model
shifts remains a largely unexplored research challenge at this point
\protect\hyperlink{ref-verma2020counterfactual}{{[}12{]}}. We have been
able to identify only one recent work that considers the implications of
\textbf{exogenous} domain and model shifts in the context of AR
\protect\hyperlink{ref-upadhyay2021towards}{{[}15{]}}. Exogenous shifts
are strictly of external origin. For example, they might stem from data
correction, temporal shifts or geospatial changes
\protect\hyperlink{ref-upadhyay2021towards}{{[}15{]}}. The authors of
\protect\hyperlink{ref-upadhyay2021towards}{{[}15{]}} propose framework
for algorithmic recourse (ROAR) that evidently improves robustness to
such exogenous shifts.

As mentioned earlier, research has so far also generally focused on
generating counterfactuals for single individuals or instances. We have
been able to identify only one existing work that investigates black-box
model behavior towards a group of individuals
\protect\hyperlink{ref-carrizosa2021generating}{{[}16{]}}. The authors
propose an optimization framework that generates collective
counterfactuals. We provide a motivation for doing so from the
perspective of endogenous macrodynamics of algorithmic recourse.

@@

In the context of algorithmic recourse, domain and model shifts were
first brought up by the authors behind ROAR
\protect\hyperlink{ref-upadhyay2021towards}{{[}15{]}}. In their work
they refer to model shifts as simply any perturbation \(\Delta\) to the
parameters of the model in question: \(M\). While this also sets the
baseline for our analysis here, it is worth noting that in
\protect\hyperlink{ref-upadhyay2021towards}{{[}15{]}} these
perturbations are mechanically introduced. In contrast we are interested
in quantifying model shifts that arise endogenously as part of a dynamic
recourse process. In addition to quantifying the magnitude of shifts
\(\Delta\), we aim to also analyse the characteristics of changes to the
model, such as the position of the decision boundary and the overall
decisiveness of the model. We have not been able to identify previous
work on this topic.

@@

Despite the large and growing number of approaches to counterfactual
search there have been surprisingly few benchmark studies that compare
different methodologies. This may be partially due to limited software
availability in this space. Recent work has started to address this gap:
firstly, \protect\hyperlink{ref-de2021framework}{{[}23{]}} run a large
benchmarking study using different algorithmic aproaches and numerous
tabular datasets; secondly,
\protect\hyperlink{ref-pawelczyk2021carla}{{[}4{]}} introduce a Python
framework that can be used to apply and benchmark different
methodologies; finally,
\protect\hyperlink{ref-altmeyer2022CounterfactualExplanations}{{[}6{]}}
provides an extensible, fast and language-agnostic implementation in
Julia. Since the experiments presented here involve extensive
simulations, we have relied on and extended the Julia implementation due
to the associated performance benefits.

@@

In the following we first set out a generalized framework for
gradient-based counterfactual search in Section~\ref{sec-method-general}
to introduce the various counterfactual generators we have chosen to use
in our experiments. We then describe the experimental setup in
Section~\ref{sec-method-2-experiment} and introduce several evaluation
metrics used to benchmark the different generators.

@@

In this work we have chosen to focus on a number of gradient-based
counterfactual generators to investigate the endogenous dynamics we
introduced in Section~\ref{sec-intro}. Gradient-based counterfactual
search is well-suited for differentiable black-box models like deep
neural networks. We can restate Equation~\ref{eq-solution} in a more
general form that encompasses most gradient-based approaches to
counterfactual search:

\begin{equation}\protect\hypertarget{eq-general}{}{
\begin{aligned}
\mathbf{s}^\prime &= \arg \min_{\mathbf{s}^\prime \in \mathcal{S}} \left\{ \sum_{k=1}^{K} {\ell(M(f(s_k^\prime)),t)}+ \lambda {h(f(s_k^\prime)) }  \right\}
\end{aligned}
}\label{eq-general}\end{equation}

Here \(\mathbf{s}^\prime=\left\{s_k^\prime\right\}_K\) is the stacked
\(K\)-dimensional array of counterfactual states and
\(f: \mathcal{S} \mapsto \mathcal{X}\) maps from the counterfactual
state space to the feature space.

@@

To explicitly address the notion that individual recourse may affect the
outcome and prospect of other individuals, we propose to extend
Equation~\ref{eq-general} as follows:

\begin{equation}\protect\hypertarget{eq-collective}{}{
\begin{aligned}
\mathbf{s}^\prime &= \arg \min_{\mathbf{s}^\prime \in \mathcal{S}}  \sum_{k=1}^{K} {\ell(M(f(s_k^\prime)),t)} \\ &+ \lambda_1 {h(f(s_k^\prime)) } + \lambda_2 {g(f(s_k^\prime))}  
\end{aligned}
}\label{eq-collective}\end{equation}

Here \(h(f(s_k^\prime))\) denotes the proxy for private costs faced by
the individual as before and \(\lambda_1\) governs to what extent that
private cost ought to be penalised. The newly introduced term
\(g(f(s_k^\prime))\) is meant to capture and address social costs
incurred by the collective of individuals in response to changes in
\(\mathbf{s}^\prime\). The trade-off between private and social costs is
determined by the ratio between \(\lambda_1\) and \(\lambda_2\). As with
individual recourse, the exact choice of \(g(\cdot)\) is not obvious,
nor do we intend to provide a definite answer in this work, if such even
exists. A straight-forward choice simply extends the baseline approach
by \protect\hyperlink{ref-wachter2017counterfactual}{{[}5{]}}: instead
of only penalizing the distance of the individuals' counterfactual to
its factual, we propose penalizing its distance to some sensible point
in the target domain, for example the sample average:
\(\bar{\mathbf{x}}\). For such a recourse objective, higher choices of
\(\lambda_2\) relative to \(\lambda_1\) will lead counterfactuals to
gravitate towards the specified point in the target domain. In the
remainder of this paper we will therefore refer to this approach as
\textbf{Gravitational} generator, when we investigate its potential
usefulness for mitigating endongenous macrodynamics\footnote{Note that
  despite the naming convention our goal here is not to provide yet
  another counterfactual generator, but merely investigate the most
  simple penalty we can think of with respect to its effectiveness.}.

\hypertarget{a-note-on-convergence}{%
\subsection{A note on convergence}\label{a-note-on-convergence}}

>>>>

\begin{document}
\maketitle
\begin{abstract}
Existing work on Counterfactual Explanations (CE) and Algorithmic
Recourse (AR) has largely been limited to the static setting and focused
on single individuals: given some estimated model the goal is to find
valid counterfactuals for individual instance that fulfill various
desiderata. The ability of such counterfactuals to handle dynamics like
data and model drift remains a largely unexplored research challenge at
this point. There has also been surprisingly little work on the related
question of how the actual implementation of recourse by one individual
may affect other individuals. Through this work we aim to close that gap
by systematizing and extending existing knowledge. We first show that
many of the existing methodologies can be collectively described by a
generalized framework. We then argue that the existing framework fails
to account for a hidden external cost of recourse, that only reveals
itself when studying the endogenous dynamics of recourse at the group
level. Through simulation experiments involving various popular
counterfactual generators and several benchmark datasets, we generate a
total XX million Counterfactual Explanations and study the resulting
domain and model shifts. We find that the induced shifts are substantial
enough to likely impede the applicability Algorithmic Recourse in
practice. Fortunately, we find various potential mitigation strategies
that can be used in combination with existing approaches. Our simulation
framework for studying recourse dynamics is fast and open-sourced.
\end{abstract}
\ifdefined\Shaded\renewenvironment{Shaded}{\begin{tcolorbox}[sharp corners, borderline west={3pt}{0pt}{shadecolor}, boxrule=0pt, enhanced, frame hidden, breakable, interior hidden]}{\end{tcolorbox}}\fi

@@

Models or algorithms that fall into this category are typically referred
to \textbf{black-box} models. Despite their shortcomings, black-box
models have grown in popularity in recent years and have at times
created undesirable societal outcomes
\protect\hyperlink{ref-o2016weapons}{{[}2{]}}. The scientific community
has tackled this issue from two different angles: while some have
appealed for a strict focus on inherently interpretable models
\protect\hyperlink{ref-rudin2019stop}{{[}3{]}}, others have investigated
different ways to explain the behaviour of black-box models. These two
sub-domains can be broadly referred to as \textbf{interpretable AI} and
\textbf{explainable AI} (XAI), respectively.

Among the approaches to XAI that have recently grown in popularity are
\textbf{Counterfactual Explanations} (CE). They explain how inputs into
a model need to change for it to produce different outputs.
Counterfactual Explanations that involve realistic and actionable
changes can be used for the purpose of \textbf{Algorithmic Recourse}
(AR) to help individuals who face adverse outcomes. An example relevant
to the Social Sciences is consumer credit: in this context, AR can be
used to guide individuals in improving their creditworthiness, should
they have previously been denied access to credit based on an automated
decision-making system. A meaningful recourse recommendation for a
denied applicant could be: \emph{``If your net savings rate had been
10\% of your monthly income instead of the actual 8\%, your application
would have been successful. See if you can temporarily cut down on
consumption.''} In the remainder of this paper we will use both
terminologies - recourse and counterfactual - interchangeably to refer
to situations where counterfactuals are generated with the intent to
provide individual recourse.

Existing work in this field has largely worked in a static setting:
various approaches have been proposed to generate counterfactuals for a
given individual that is subject to some pre-trained model. More recent
work has compared different approaches within this static setting
\protect\hyperlink{ref-pawelczyk2021carla}{{[}4{]}}. In this work, we go
one step further and ask ourselves: what happens if recourse is provided
and implemented repeatedly? What types of dynamics are introduced and
how do different counterfactual generators compare in this context?

Research on Algorithmic Recourse has also so far typically addressed the
issue from the perspective of one single individual and has indeed been
referred to as \textbf{individual recourse} in some places. Arguably
though, most real-world applications that warrant Algorithmic Recourse
involve potentially large groups of individuals typically competing for
scarce resources. Our work demonstrates that in such scenarios, choices
made by or for one single individual are likely to affect the broader
collective of individuals in ways that current approaches to AR fail to
account for. More specifically, we argue that a strict focus on
minimizing the private costs faced by individuals may be too narrow an
objective.

@@

To the best of our knowledge this is the first work investigating
endogenous macrodynamics in AR. Our contributions to the state of
knowledge are as follows: firstly, we posit a compelling argument that
calls for a novel perspective on Algorithmic Recourse extending our
focus from single individuals to groups. Secondly, we introduce an
experimental framework extending previous work by
\protect\hyperlink{ref-altmeyer2022CounterfactualExplanations}{{[}6{]}},
which enables us to study macrodynamics of Algorithmic Recourse through
simulations that can be fully parallelized. Thirdly, we use this
framework to provide a first in-depth analysis of endogenous recourse
dynamics induced by various popular counterfactual generators including
\protect\hyperlink{ref-wachter2017counterfactual}{{[}5{]}},
\protect\hyperlink{ref-schut2021generating}{{[}7{]}},
\protect\hyperlink{ref-joshi2019towards}{{[}8{]}},
\protect\hyperlink{ref-mothilal2020explaining}{{[}9{]}} and
\protect\hyperlink{ref-antoran2020getting}{{[}10{]}}. To this end we
propose a number of novel evaluation metrics that can be used to
quantify and benchmark the macrodynamics introduced by the different
generators. Finally, we also discuss what drives endogenous dynamics and
propose strategies to mitigate them.

The remainder of the paper is structured as follows:
Section~\ref{sec-related} places our work in the broader context of
related literature. Section~\ref{sec-method} posits a generalized
framework for gradient-based Algorithmic Recourse and introduces the
notion of hidden external costs. Section~\ref{sec-method-2} sets out our
experimental framework for modeling endogenous macrodynamics in AR.
Section~\ref{sec-empirical} presents our experimental results for
synthetic and real-world datasets. We also present evidence for the
effectiveness of various proposed mitigation strategies in that section.
Our findings are then discussed in the broader context of the literature
in Section~\ref{sec-discussion}. We also point to some of the
limitations of our work as well as avenues for future research in
Section~\ref{sec-limit}. Finally, Section~\ref{sec-conclusion}
concludes.

@@

In this Section we provide a review of the relevant literature. First,
Section~\ref{sec-related-recourse} discusses the existing research
within the domain of Counterfactual Explanations and Algorithmic
Recourse. Then, Section~\ref{sec-related-shifts} presents some of the
previous work on the measurement of dataset and model shifts.

@@

A framework for Counterfactual Explanations was first proposed in 2017
by \protect\hyperlink{ref-wachter2017counterfactual}{{[}5{]}} and has
served as the baseline for most methodologies that have been proposed
since then. Let \(M: \mathcal{X} \mapsto \mathcal{Y}\) denote some
pre-trained model that maps from inputs \(X \in \mathcal{X}\) to outputs
\(Y \in \mathcal{Y}\). Then we are interested in minimizing the
cost\footnote{Equivalently, others have referred to this quantity as
  \emph{complexity} or simply \emph{distance}} \(H=h(x\prime)\) incurred
by individual \(x\) when moving to a counterfactual state \(x\prime\)
such that the predicted outcome \(M(x\prime)\) corresponds to some
target outcome \(y^*\):

\begin{equation}\protect\hypertarget{eq-obj}{}{
\min_{x\prime \in \mathcal{x}} h(x\prime) \ \ \ \mbox{s. t.} \ \ \ M(x\prime) = t
}\label{eq-obj}\end{equation}

@@

\begin{equation}\protect\hypertarget{eq-solution}{}{
x\prime = \arg \min_{x\prime}  \ell(M(x\prime),y^*) + \lambda h(x\prime)
}\label{eq-solution}\end{equation}

In the baseline work
\protect\hyperlink{ref-wachter2017counterfactual}{{[}5{]}}, the cost
function \(h: \mathcal{X} \mapsto \mathbb{R}\) is proxied by some
distance metric based on the simple intuition that perturbations of
\(x\) are costly to the individual. For models that are differentiable
and produce smooth predictions, Equation~\ref{eq-solution} can be solved
through gradient descent. This summarizes the approach followed in
\protect\hyperlink{ref-wachter2017counterfactual}{{[}5{]}} which we
shall refer to simply as \textbf{Wachter} - the name of the first author
- in the remainder of this paper.

Many approaches for the generation of Algorithmic Recourse have been
described in the literature since 2017. An October 2020 survey by Karimi
et al.~layed out 60 algorithms that have been proposed since 2014
\protect\hyperlink{ref-karimi2020survey}{{[}11{]}}. Another survey
published around the same time by Verma et al.~described 29 algorithms
\protect\hyperlink{ref-verma2020counterfactual}{{[}12{]}}. Different
approaches vary primarily in terms of the objective functions they
impose, how they optimize said objective (from brute force through
gradient-based approaches to graph traversal algorithms), and how the
ensure that certain requirements for CE are met. Regarding the latter,
the literature has produced an extensive list of desiderata each
addressing different needs. To name but a few, we are interested in
generating counterfactuals that close
\protect\hyperlink{ref-wachter2017counterfactual}{{[}5{]}}, actionable
\protect\hyperlink{ref-ustun2019actionable}{{[}13{]}}, realistic
\protect\hyperlink{ref-schut2021generating}{{[}7{]}}, sparse, diverse
\protect\hyperlink{ref-mothilal2020explaining}{{[}9{]}} and if possible
causally founded
\protect\hyperlink{ref-karimi2021algorithmic}{{[}14{]}}.

Efforts so far have largely been directed at improving the quality of
Counterfactual Explanations within a static context: given some
pre-trained classifier \(M: \mathcal{X} \mapsto \mathcal{Y}\), we are
interested in generating one or multiple meaningful Counterfactual
Explanations for some individual characterized by \(x\). The ability of
Counterfactual Explanations to handle dynamics like data and model
shifts remains a largely unexplored research challenge at this point
\protect\hyperlink{ref-verma2020counterfactual}{{[}12{]}}. We have been
able to identify only one recent work that considers the implications of
\textbf{exogenous} domain and model shifts in the context of AR
\protect\hyperlink{ref-upadhyay2021towards}{{[}15{]}}. Exogenous shifts
are strictly of external origin. For example, they might stem from data
correction, temporal shifts or geospatial changes
\protect\hyperlink{ref-upadhyay2021towards}{{[}15{]}}. The authors of
\protect\hyperlink{ref-upadhyay2021towards}{{[}15{]}} propose ROAR - a
framework for Algorithmic Recourse that evidently improves robustness to
such exogenous shifts.

As mentioned earlier, research has so far also generally focused on
generating counterfactuals for single individuals or instances. We have
been able to identify only one existing work that investigates black-box
model behavior towards a group of individuals
\protect\hyperlink{ref-carrizosa2021generating}{{[}16{]}}. The authors
propose an optimization framework that generates collective
counterfactuals. We provide a motivation for doing so from the
perspective of endogenous macrodynamics of Algorithmic Recourse.

@@

In the context of Algorithmic Recourse, domain and model shifts were
first brought up by the authors behind ROAR
\protect\hyperlink{ref-upadhyay2021towards}{{[}15{]}}. In their work
they refer to model shifts as simply any perturbation \(\Delta\) to the
parameters of the model in question: \(M\). While this also sets the
baseline for our analysis here, it is worth noting that in
\protect\hyperlink{ref-upadhyay2021towards}{{[}15{]}} these
perturbations are mechanically introduced. In contrast we are interested
in quantifying model shifts that arise endogenously as part of a dynamic
recourse process. In addition to quantifying the magnitude of shifts
\(\Delta\), we aim to also analyse the characteristics of changes to the
model, such as the position of the decision boundary and the overall
decisiveness of the model. We have not been able to identify previous
work on this topic.

@@

Despite the large and growing number of approaches to counterfactual
search, there have been surprisingly few benchmark studies that compare
different methodologies. This may be partially due to limited software
availability in this space. Recent work has started to address this gap:
firstly, \protect\hyperlink{ref-de2021framework}{{[}23{]}} run a large
benchmarking study using different algorithmic aproaches and numerous
tabular datasets; secondly,
\protect\hyperlink{ref-pawelczyk2021carla}{{[}4{]}} introduce a Python
framework - CARLA - that can be used to apply and benchmark different
methodologies; finally, \texttt{CounterfactualExplanations.jl}
\protect\hyperlink{ref-altmeyer2022CounterfactualExplanations}{{[}6{]}}
provides an extensible, fast and language-agnostic implementation in
Julia. Since the experiments presented here involve extensive
simulations, we have relied on and extended the Julia implementation due
to the associated performance benefits. In particular, we have built a
framework on top of \texttt{CounterfactualExplanations.jl} that extends
the functionality from static benchmarks to simulation experiments:
\texttt{AlgorithmicRecourseDynamics.jl}\footnote{The package is
  available from \ldots{}}. The core concepts implemented in that
package reflect what is presented in section Section~\ref{sec-method-2}
of this paper.

@@

In this section we first set out a generalized framework for
gradient-based counterfactual search in Section~\ref{sec-method-general}
that encapsulates the various individual recourse methods we have chosen
to use in our experiments. We then introduce the notion of a hidden
external cost in algorithmic recourse and extend the existing framework
to explicitly address this cost in the counterfactual search objective.

@@

We have chosen to focus on gradient-based counterfactual search for two
reasons: firstly, they can be seen as direct descendants of our baseline
method - Wachter; secondly, gradient-based search is particularly
well-suited for differentiable black-box models like deep neural
networks, which we focus on in this work. In particular, we include
include the following generators in our simulation experiments below:
\textbf{REVISE} \protect\hyperlink{ref-joshi2019towards}{{[}8{]}},
\textbf{CLUE} \protect\hyperlink{ref-antoran2020getting}{{[}10{]}},
\textbf{DiCE} \protect\hyperlink{ref-mothilal2020explaining}{{[}9{]}}
and a greedy approach that relies on probabilistic models
\protect\hyperlink{ref-schut2021generating}{{[}7{]}}. Our motivation for
including these different generators in our analysis, is that they all
offer slightly different approaches to generate meaningful
counterfactuals for differentiable black-box models. We hypothesize that
generating more \textbf{meaningful} counterfactuals should mitigate the
endogenous dynamics illustrated in Figure~\ref{fig-poc} in
Section~\ref{sec-intro}. This intuition stems from the underlying idea
that more meaningful counterfactuals are generated by the same or at
least a very similar data generating process as the training data. All
else equal, counterfactuals that fulfill this basic requirement should
be less prone to trigger domain and model shifts.

As we will see next, all of them can be described by the following
generalized form of Equation~\ref{eq-solution}:

\begin{equation}\protect\hypertarget{eq-general}{}{
\begin{aligned}
\mathbf{s}^\prime &= \arg \min_{\mathbf{s}^\prime \in \mathcal{S}} \left\{ \sum_{k=1}^{K} {\ell(M(f(s_k^\prime)),y^*)}+ \lambda {h(f(s_k^\prime)) }  \right\}
\end{aligned}
}\label{eq-general}\end{equation}

Here \(\mathbf{s}^\prime=\left\{s_k^\prime\right\}_K\) is the stacked
\(K\)-dimensional array of counterfactual states and
\(f: \mathcal{S} \mapsto \mathcal{X}\) maps from the counterfactual
state space to the feature space. In Wachter, the state space is the
feature space: \(f\) is just the identity function and the number of
counterfactuals \(K\) is equal to one. Both REVISE and CLUE search
counterfactuals in some latent embedding \(S \subset \mathcal{S}\)
instead of the feature space directly. The latent embedding is learned
by a separate generative model that is tasked with learning the data
generating process (DGP) of \(X\). In this case \(f\) in
Equation~\ref{eq-general} corresponds to the decoder part of the
generative model, in other words the function that maps back from the
latent embedding to the feature space. Provided the generative model is
well-specified, traversing the latent embedding typically results in
realistic and plausible counterfactuals, because they are implicitly
generated by the (learned) DGP
\protect\hyperlink{ref-joshi2019towards}{{[}8{]}}.

CLUE distinguishes itself from REVISE and other counterfactual
generators in that it aims to minimize the predictive uncertainty of the
model in question, \(M\). To quantify predictive uncertainty the authors
rely on entropy estimates for probabilistic models. The approach
proposed by \protect\hyperlink{ref-schut2021generating}{{[}7{]}}, which
we shall refer to as \textbf{Greedy}, also works with the subclass of
models \(\tilde{\mathcal{M}}\subset\mathcal{M}\) that can produce
predictive uncertainty estimates. The authors show that in this setting
the cost function \(h(\cdot)\) in Equation~\ref{eq-general} is redundant
and meaningful counterfactuals can be generated in a fast and efficient
manner through a modified Jacobian-based Saliency Map Attack (JSMA).
Schut et al. \protect\hyperlink{ref-schut2021generating}{{[}7{]}} also
show that by maximizing the predicted probability of \(x^\prime\) being
assigned to target class \(y^*\) we also implicitly minimize predictive
entropy - as in CLUE. In that sense, CLUE can be seen as equivalent to
REVISE in the Bayesian context and we shall therefore refer to both
approaches collectively as \textbf{Latent Space} generators\footnote{In
  fact, there are a number of other recently proposed approaches to
  counterfactual search that also broadly fall in this same category.
  They largely differ with respect to the chosen generative model: for
  example, the generator proposed by
  \protect\hyperlink{ref-dombrowski2021diffeomorphic}{{[}24{]}} relies
  on normalizing flows.}.

Finally, DiCE \protect\hyperlink{ref-mothilal2020explaining}{{[}9{]}}
distinguishes itself from all other generators considered here in that
it aims to generate a diverse set of \(K>1\) counterfactuals. Wachter et
al. \protect\hyperlink{ref-wachter2017counterfactual}{{[}5{]}} show that
diverse outcomes can in principal be achieved simply rerunning
counterfactual search multiple times using stochastic gradient descent
(or by randomly initializing the counterfactual). In
\protect\hyperlink{ref-mothilal2020explaining}{{[}9{]}} diversity is
explicitly proxied via Determinantal Point Processes (DDP): the authors
simply introduce DDP as a component of the cost function
\(h(\mathbf{s}^\prime)\) and therebt produce counterfactuals
\(s_1, ... , s_K\) that look as different from each other as possible.
The implementation of DiCE in the our library of choice -
\texttt{CounterfactualExplanations.jl} - uses that exact approach. It is
worth noting that for \(k=1\), DiCE reduces to Wachter since the DDP is
constant and therefore does not affect the objective function
Equation~\ref{eq-general}.

@@

All of the different approaches introduced above tackle the problem of
Algorithmic Recourse from the perspective of one single
individual\footnote{DiCE recognizes that different individuals may have
  different objective functions, but it does not address the
  interdependencies between different individuals.}. To explicitly
address the issue that individual recourse may affect the outcome and
prospect of other individuals, we propose to extend
Equation~\ref{eq-general} as follows:

\begin{equation}\protect\hypertarget{eq-collective}{}{
\begin{aligned}
\mathbf{s}^\prime &= \arg \min_{\mathbf{s}^\prime \in \mathcal{S}}  \sum_{k=1}^{K} {\ell(M(f(s_k^\prime)),y^*)} \\ &+ \lambda_1 {h(f(s_k^\prime)) } + \lambda_2 {g(f(s_k^\prime))}  
\end{aligned}
}\label{eq-collective}\end{equation}

Here \(h(f(s_k^\prime))\) denotes the proxy for private costs faced by
the individual as before and \(\lambda_1\) governs to what extent that
private cost ought to be penalized. The newly introduced term
\(g(f(s_k^\prime))\) is meant to capture and address external costs
incurred by the collective of individuals in response to changes in
\(\mathbf{s}^\prime\). The underlying concept of private and external
costs is borrowed from Economics and well-established in that field:
when the decisions or actions by some individual market participant
generate external costs, then the market is said to suffer from negative
externalities and considered inefficient
\protect\hyperlink{ref-pindyck2014microeconomics}{{[}25{]}}. We think
that this concept describes the endogenous dynamics of algorithmic
recourse oberserved here very well. As with individual recourse, the
exact choice of \(g(\cdot)\) is not obvious, nor do we intend to provide
a definite answer in this work, if such even exists. That being said, we
do propose a few potential mitigation strategies in
Section~\ref{sec-empirical-2-mitigate}.

In the following we describe the framework we propose for modeling and
analysing endogenous macrodynamics in Algorithmic Recourse. We first
describe the basic simulations that were generated to produce the
findings in this work and also constitute the core of
\texttt{AlgorithmicRecourseDynamics.jl} - the Julia package we
introduced earlier. The remainder of this section then introduces
various evaluation metrics that can be used to benchmark different
counterfactual generators with respect to how they perform in the
dynamic setting.

The dynamics illustrated in Figure~\ref{fig-poc} in
Section~\ref{sec-intro} were generated through a simple experiment that
aims to simulate the process of Algorithmic Recourse in practice. We
begin in the static setting at time time \(t=0\): given some pre-trained
classifier \(M\) we generate recourse for a random batch of \(B\)
individuals in the non-target class. Note that we focus our attention on
classification problems, since classification poses the most common
practical use-case for Algorithmic Recourse. In order to simulate the
dynamic process, we suppose that the model \(M\) is retrained following
the actual implementation of recourse in time \(t=0\). Following the
update to the model, we assume that at time \(t=1\) recourse is
generated for yet another random subset of individuals in the non-target
class. This process is repeated for a number of time periods \(T\). To
get a clean read on endogenous dynamics we keep the total population of
samples closed: we allow existing samples to move from factual to
counterfactual states, but do not allow any entirely new samples to
enter the population. The experimental setup is summarized in Algorithm
\ref{algo-experiment}

\begin{algorithm}
\caption{Simulation Experiment}\label{algo-experiment}
\begin{algorithmic}[1]
\Procedure{Experiment}{$M,D,G$}
\State $t\gets 0$
\While{$t<T$}
\State $D_B \subset D$
\State $D_B\gets G(D_B)$ \Comment{Generate counterfactuals.}
\State $M\gets M(D)$ \Comment{Retrain model.}
\EndWhile
\State \textbf{return} $M,D$
\EndProcedure
\end{algorithmic}
\end{algorithm}
